---
title: "covid_valid_am"
author: "Alisa Morshneva"
date: "28/11/2021"
output: html_document
---

# External validation of the prognostic score for the regular monitoring of COVID-19 patients condition

Settings:
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message=FALSE,
                      error=FALSE,
                      warning=FALSE)

library(tidyr)
require(lubridate)
library(ggplot2)
library(stringr)
library(zoo)
library(flextable)
library(dplyr)
library(readxl)


```

## Read data
```{r}
transferuri_raw <- read_excel("COVID_PROJECT_UPDATE_16.11.2021.xlsx")
transferuri_raw <- read_excel("COVID_PROJECT_UPDATE_16.11.2021.xlsx")
#analysis_p <- read.csv("data/analize_planice_new.csv")
#analysis_u <- read.csv("data/analize_urgenta_new.csv")
#info_raw <- read.csv("data/total.csv")
```

## Parse data
```{r}
# parse date and time:
transferuri <-transferuri_raw %>% 
  separate(StartDate, c("StartDate", "StartTime"), sep = " ") %>%
  separate(EndDate, c("EndDate", "EndTime"), sep = " ")
# set date format:
transferuri$StartDate <- as.Date(as.character(transferuri$StartDate), format="%d/%m/%Y")
transferuri$EndDate <- as.Date(as.character(transferuri$EndDate), format="%d/%m/%Y")
# calculate days of hospitalization and days before an outcome 
transferuri$hospit_days <- transferuri$EndDate - transferuri$StartDate
hosp_days <- transferuri %>%                                       
  group_by(id) %>%                         
  summarise(hospit_days_total = sum(hospit_days)) 

transferuri <-  merge(transferuri, hosp_days, by="id")
transferuri$OutcomeDate <- transferuri$StartDate + transferuri$hospit_days_total 

# unify data for the integration
transfer <- select(transferuri, c(id, StartDate, hospit_days_total, OutcomeDate))
transfer <- transfer[row.names(unique(transfer[,c("id", "hospit_days_total")])),]

colnames(analysis_p) <-  c("id", "AnalysisDate", "analiza", "rezultat")
colnames(analysis_u) <- c("id", "AnalysisDate", "analiza", "rezultat")
analysis <- rbind(analysis_p, analysis_u)
# set date format:
analysis$AnalysisDate <- as.Date(as.character(analysis$AnalysisDate), format="%d/%m/%Y")

# prepare info table
colnames(info_raw)[5] <- "Age"
colnames(info_raw)[11] <- "Outcome"
info <- info_raw[,c("id", "Age", "sex", "cicode","Outcome")]
```
## Merge data
```{r}
data_long_raw <- merge(transfer, analysis, by=c("id"))
# add outcomes
data_long_raw <- merge(data_long_raw, info, by=c("id"))
# count initial number of unique patients
length(unique(data_long_raw$id))
```
## Filter data
```{r}
# filter data by diagnosis (only respiratory)
data_long <- filter(data_long_raw, cicode %in% c("J128", "J129", "J168", "J188", "J189", "J208", "B342", "B349", "B972"))
# remove patients transfered to other hospitals 
data_long <- filter(data_long, Outcome != 'transfer inter-spitalicesc')
# set outcomes (0 - discharged, 1 - deceased)
data_long$Outcome <- as.factor(ifelse(substring(tolower(data_long$Outcome),1,1)=='e', 0, 1))
# filter data by outcome time (less than 3 days in hospital):
data_long <- filter(data_long, hospit_days_total > 2)
# count number of unique patients
length(unique(data_long$id))
```
## Parse data
```{r}
# prepare data for conversion to long format
data_long$rezultat <- gsub('[_><]', '', data_long$rezultat)
data_long$rezultat <- gsub('min', '', data_long$rezultat)
data_long$rezultat <- trimws(data_long$rezultat, which = c("both"), whitespace = "[ \t\r\n]")
data_long <- filter(data_long, !str_detect(rezultat, '[A-z+/]'))
data_long$rezultat <- as.numeric(as.character(data_long$rezultat))
#count NA
table(is.na(data_long$rezultat))
```
## Name and format corrections
```{r}
data_long_upd <- data_long

data_long_upd$analiza <- gsub('D-dimeri', 'D-Dimer', data_long_upd$analiza)
data_long_upd$rezultat <- ifelse(data_long_upd$analiza == 'D-Dimer', data_long_upd$rezultat * 1000, data_long_upd$rezultat)
data_long_upd$analiza <- ifelse(data_long_upd$analiza %in% c('Timpul de tromboplastina partial activa','TTPA-Timpul de tromboplastina partial active','Timpul de tromboplastina  partial active  (TTPA)') , "APTT", data_long_upd$analiza)
data_long_upd$analiza <- ifelse(data_long_upd$analiza %in% c('Glucose','Glucoza','Dozarea glucozei','Dozarea glucozei in sange') , "Glucose", data_long_upd$analiza)
data_long_upd$analiza <- ifelse(data_long_upd$analiza %in% c('Dozarea ureei','Ureea','Uree') , "Urea", data_long_upd$analiza)
data_long_upd$analiza <- ifelse(data_long_upd$analiza %in% c('Leucocites','Numaratoarea leucocitelor') , "WBC", data_long_upd$analiza)
data_long_upd$analiza <- tolower(data_long_upd$analiza)
data_long_upd$analiza <- gsub('  ', ' ', data_long_upd$analiza)
data_long_upd$analiza <- gsub(':', '', data_long_upd$analiza)
data_long_upd$analiza <- gsub('+', '', data_long_upd$analiza, fixed = TRUE)
data_long_upd$analiza <- gsub('ureea', 'uree', data_long_upd$analiza)
data_long_upd$analiza <- gsub('uree', 'urea', data_long_upd$analiza)
data_long_upd$analiza <- gsub('determinarea proteinei c-reactive', 'crp', data_long_upd$analiza) 
data_long_upd$analiza <- gsub('dozarea proteinei totale', 'protein_total', data_long_upd$analiza) 
data_long_upd$analiza <- gsub('total protein', 'protein_total', data_long_upd$analiza) 
data_long_upd$analiza <- gsub('dozarea hemoglobinei', 'hg', data_long_upd$analiza) 
data_long_upd$analiza <- gsub('singerare', 'sangerare', data_long_upd$analiza)
data_long_upd$analiza <- gsub('clor', 'cl', data_long_upd$analiza)
data_long_upd$analiza <- gsub('figrinogen', 'fibrinogen', data_long_upd$analiza, fixed=TRUE)
data_long_upd$analiza <- gsub('fibrinogenul', 'fibrinogen', data_long_upd$analiza, fixed=TRUE) 
data_long_upd$analiza <- gsub('ca2', 'calciu', data_long_upd$analiza, fixed=TRUE) 
data_long_upd$analiza <- gsub(' o (aslo)', 'o', data_long_upd$analiza, fixed=TRUE) 
data_long_upd$analiza <- gsub('densitatea', 'densitate', data_long_upd$analiza, fixed=TRUE) 
data_long_upd$analiza <- gsub('centriguare', 'centrifugare', data_long_upd$analiza, fixed=TRUE) 
data_long_upd$analiza <- trimws(data_long_upd$analiza, which = c("both"), whitespace = "[ \t\r\n]")
```

## Convert the format (long to wide) and fill empty values with previous results (locf)
```{r}
data_long_to_wide <- data_long_upd %>%
  group_by(id, StartDate, OutcomeDate, AnalysisDate, Outcome, analiza) %>% 
  dplyr::summarise(rezultat_med = median(rezultat, na.rm = TRUE))

data_full <-  pivot_wider(data_long_to_wide, names_from = "analiza", values_from = "rezultat_med") %>%
  group_by(id) %>% arrange(id,AnalysisDate) %>%  mutate_all(funs(na.locf(., na.rm = FALSE)))

```

## Set formats and sort columns
```{r}
data_full$DaysBfOutcome <- as.character(data_full$OutcomeDate - data_full$AnalysisDate)
data_full$DayOfHosp <- as.character(data_full$AnalysisDate - data_full$StartDate)
data_full$limphocite_perc <- (data_full$limfocite/100)*data_full$wbc
data_full <- select(data_full,-c(StartDate,OutcomeDate))
data_full$id <- as.factor(data_full$id)

# filter by outcome time (0-21 days)
data <- data_full %>% 
  select(sapply(., class) %>% .[order(match(., c('factor', 'date', 'character','integer')))] %>% names) %>%
  mutate(DaysBfOutcome = as.numeric(DaysBfOutcome)) %>%
  filter(DaysBfOutcome >= 0 & DaysBfOutcome < 21) %>%
  mutate(DaysBfOutcome = factor(DaysBfOutcome,
                                levels = 21:0,
                                labels = 21:0,
                                ordered = TRUE))
# write reformatted data to file
write.csv(data, file = "data_all_analyzes.csv")
```
## Select 9 features:
```{r}
data_9 <- select(data, c(id, Outcome, DaysBfOutcome, DayOfHosp, AnalysisDate, 'd-dimer', urea, limphocite_perc, crp, protein_total, aptt, wbc, hg, glucose)) 
data_9
```

# Draw boxplots 
```{r}
ggplot(data %>% filter(`protein_total` < 200), aes(x = DaysBfOutcome, y = protein_total, fill = Outcome)) + 
  geom_boxplot() 

ggplot(data, aes(x = DaysBfOutcome, y = `urea`, fill = Outcome)) + 
  geom_boxplot()

ggplot(data, aes(x = DaysBfOutcome, y = `d-dimer`, fill = Outcome)) + 
  geom_boxplot() + scale_y_log10()

ggplot(data, aes(x = DaysBfOutcome, y = `wbc`, fill = Outcome)) + 
  geom_boxplot() + scale_y_log10() 


ggplot(data, aes(x = DaysBfOutcome, y = limphocite_perc , fill = Outcome)) + 
  geom_boxplot() + ylim(0,10)
```
## Count prognostic score:
```{r}
score_9 <- data_9
score_9$score <- 0
score_9$score  <- ifelse(score_9$'d-dimer' > 2149 & !is.na(score_9$'d-dimer'), score_9$score + 4, score_9$score)
score_9$score  <- ifelse(score_9$urea > 11 & !is.na(score_9$urea), score_9$score + 5, score_9$score)
score_9$score  <- ifelse(score_9$limphocite_perc < 0.7 & !is.na(score_9$limphocite_perc), score_9$score + 3, score_9$score)
score_9$score  <- ifelse(score_9$crp > 146 & !is.na(score_9$crp), score_9$score + 3, score_9$score)
score_9$score  <- ifelse(score_9$protein_total < 61 & !is.na(score_9$protein_total), score_9$score + 6, score_9$score)
score_9$score  <- ifelse(score_9$aptt > 42 & !is.na(score_9$aptt), score_9$score + 4, score_9$score)
score_9$score  <- ifelse(score_9$wbc > 13.5 & !is.na(score_9$wbc), score_9$score + 4, score_9$score)
score_9$score  <- ifelse(score_9$hg < 115 & !is.na(score_9$hg), score_9$score + 3, score_9$score)
score_9$score  <- ifelse(score_9$glucose > 9 & !is.na(score_9$glucose), score_9$score + 4, score_9$score)
score_9
write.csv(score_9,file = "score_9.csv")
```

## Plot the score dynamics
```{r}
p <- ggplot(score_9, aes(x = DaysBfOutcome, y = score, fill = Outcome)) + 
  geom_boxplot()
p
```
```{r}
summarySE <- function(data=NULL, measurevar, groupvars=NULL, na.rm=FALSE,
                      conf.interval=.95, .drop=TRUE) {
    library(plyr)

    # New version of length which can handle NA's: if na.rm==T, don't count them
    length2 <- function (x, na.rm=FALSE) {
        if (na.rm) sum(!is.na(x))
        else       length(x)
    }

    # This does the summary. For each group's data frame, return a vector with
    # N, mean, and sd
    datac <- ddply(data, groupvars, .drop=.drop,
      .fun = function(xx, col) {
        c(N    = length2(xx[[col]], na.rm=na.rm),
          mean = mean   (xx[[col]], na.rm=na.rm),
          sd   = sd     (xx[[col]], na.rm=na.rm)
        )
      },
      measurevar
    )

    # Rename the "mean" column    
    datac <- rename(datac, c("mean" = measurevar))

    datac$se <- datac$sd / sqrt(datac$N)  # Calculate standard error of the mean

    # Confidence interval multiplier for standard error
    # Calculate t-statistic for confidence interval: 
    # e.g., if conf.interval is .95, use .975 (above/below), and use df=N-1
    ciMult <- qt(conf.interval/2 + .5, datac$N-1)
    datac$ci <- datac$se * ciMult

    return(datac)
}

sum <- summarySE(score_9, measurevar="score", groupvars=c("DaysBfOutcome","Outcome"))

pd <- position_dodge(0.1) # move them .05 to the left and right

p <-ggplot(sum, aes(x=DaysBfOutcome, y=score, colour=Outcome)) + 
    geom_errorbar(aes(ymin=score-ci, ymax=score+ci), width=.1, position=pd) +
    geom_line(position=pd) +
    geom_point(position=pd) + theme_bw() + ylab("Average score + 95% CI") + xlab("Day before outcome") +  scale_color_manual(labels=c("Discharge","Death"), values = c("#00BFC4", "#F8766B"))
p
```

## Expected death/discharge odds:
```{r}
# count max_score
score_9_max <- score_9 %>% group_by(id, Outcome) %>% summarise(max_score = max(score, na.rm = TRUE)) %>% filter(!is.na(max_score))

# logistic regression
glm_outcome <- glm(Outcome ~ max_score, family=binomial, data = score_9_max)
#summary(glm_outcome)

# outcome prediction
max_score <- seq(1,36, by = 1)
scores <- as.data.frame(max_score)

predictions <- predict.glm(glm_outcome, scores, type="response", se.fit = TRUE)

# count confidence intervals
scores$outcome_pr <- predictions$fit
ci <- 1.96 * predictions$se.fit
scores$ci_low_pr <- predictions$fit - ci
scores$ci_up_pr <- predictions$fit + ci

# count odds (odd = death probability/discharge probability)
scores$odd <- round(scores$outcome_pr/(1- scores$outcome_pr), 5)
scores$ci_low_odd <- round(scores$ci_low_pr/(1- scores$ci_low_pr), 5)
scores$ci_up_odd <- round(scores$ci_up_pr/(1-scores$ci_up_pr), 5)
scores$max_score <- as.factor(scores$max_score)

# plot odds
pd <- position_dodge(0.1) # move them .05 to the left and right
ggplot(scores, aes(x=max_score, y=odd)) + 
    geom_errorbar(aes(ymin=ci_low_odd, ymax=ci_up_odd), width=.1, position=pd) +
    geom_line(position=pd) +
    geom_point(position=pd) + theme_bw() + scale_y_log10(limits = c(0.001, 200), breaks=c(0.001,0.01,0.1,1,10,100), labels = c("1/1000", "1/100", "1/10", "1","10","100")) 
```


# Set score grades (patient individual risk)
```{r}
thresholds <- c("<4", "[4, 8)", "[8, 14)", "[14, 20)", ">=20")
odds <- c("< 1:100","1:100 - 1:25", "1:25 -  1:5", "1:5 - 1:1", ">1:1")
risk_grade <- c("Very Low", "Low", "Average", "High", "Very High")

data <- data.frame(thresholds, odds, risk_grade)
colnames(data) <- c("Score range", "Expected death/discharge odds", "Risk grade")
data %>% flextable(cwidth = 1.5) %>%
  theme_box() %>%
  align(align = "center", part = "all") 
```
# Plot the risk grades
```{r}
data <- filter(scores, max_score %in% c(4,8,14,20))

ggplot(data, aes(x=max_score, y=odd, color = "")) + 
    geom_errorbar(aes(ymin=ci_low_odd, ymax=ci_up_odd), width=.1, position=pd) +
    geom_line(position=pd) +
    geom_point(position=pd) + theme_bw() + scale_y_log10(breaks = c(0.01,0.04,0.2,1), ,labels = c("1/100","1/25","1/5","1")) + ylab("Odds + 95% CI") + xlab("Score grades") + theme(legend.position = "none")

```












